---
title: "Project 4- Recommender System"
author: "mohits3, usyadav2, kamlesh2"
output:
  html_document:
    df_print: paged
---


## Introduction

In this project, we have designed a recommender system which recommends movie to its Users. We have worked on below 3 usecases.

- System 1 - In this recommender system predicts the movies based on the genres selected by the User.

- System 2 - In this recommender system evaluate two most popular recommnder approaches by using performance parameter as RMSEs.
  - UBCF : (User Based collaborative Filtering)
    - parameter: normalize = 'Z-score', method = 'cosine', nn = 5 
  - SVDF: (Singular Value Decomposition Filtering) Funk SVD with gradient descend
    - parameter: normalize = 'Z-score'
  
- Shiny App - Build a user interface by shiny app demonstrating System-1 (recommendation by genre) and System-2 (User Based Colloborative Filtering) algorithm.

#### Data Description
 Dataset used in this project is MovieLens 1M dataet. The dataset contains 3 files: **movies.dat**, **ratings.dat**, **users.dat**. These files contain 1,000,209 anonymous ratings of approximately 3,900 movies made by 6,040 MovieLens users who joined MovieLens in 2000.


### Package Load

```{r, warning=FALSE, message=FALSE}
mypackages = c("recommenderlab", "ggplot2", "dplyr", "reshape2","knitr")   # required packages
tmp = setdiff(mypackages, rownames(installed.packages()))  # packages need to be installed
if (length(tmp) > 0) install.packages(tmp)
lapply(mypackages, require, character.only = TRUE)
```

### Loading Ratings, Movies and User Data

```{r, warning=FALSE, message=FALSE}
myurl = "https://liangfgithub.github.io/MovieData/"


###  Reading Ratings Data   ####
# use colClasses = 'NULL' to skip columns
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
ratings$Timestamp = NULL;
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')


###  Reading Movies Data  ###
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)
# convert accented characters
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))
# read in movie images
small_image_url = "https://liangfgithub.github.io/MovieImages/"
movies$image_url = sapply(movies$MovieID, 
                          function(x) paste0(small_image_url, x, '.jpg?raw=true'))



###    Reading User Data   ###
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
```


## System 1
### Recommendation by Genre (First Approach - Highly Popular)

In System 1 recommendation, movies are recommended to users by Genre, In first approach, we are recommending top 5 most popular thriller movies. In first recommendation, we are considering the most popular movies are those movies which are receiving the most number of ratings by movie users. Such movies may not be highly rated by them.

```{r, warning=FALSE, message=FALSE}
## Genre Based Recommendation Function by Popularity

get_recommendations_by_genre_by_popularity <- function(movies, ratings, user_genre){
  # combine rating information into movies
  movieRating = ratings %>% 
    group_by(MovieID) %>% 
    summarize(ratings_per_movie = n(), ave_ratings = round(mean(Rating), dig=3)) %>%
    inner_join(movies, by = 'MovieID')
  
  # Filter movies from selected genre
  mr_selected = movieRating[grep(user_genre, movieRating$Genres),]
  
  # recommender based on popularity
  recommendations = arrange(mr_selected, desc(ratings_per_movie))
  
  # Select Top 10 Movies
  recommendations = recommendations %>%
    select('MovieID', 'Title', 'ave_ratings', 'ratings_per_movie', 'Genres', 'Year') %>%
    top_n(5, ratings_per_movie)
  recommendations
}

## Hardcoding Genre as "Thriller"
user_genre = "Thriller"
recom_pop = get_recommendations_by_genre_by_popularity(movies, ratings, user_genre)
kable(recom_pop,format = 'markdown')
```

### Recommendation by Genre (Second Approach - Highly Rated)

In second approach, we are recommending top 5 most highly rated thriller movies. We are considering most highly rated movies are those movies which are receiving high ratings.
```{r, warning=FALSE, message=FALSE}
get_recommendations_by_genre_by_rating <- function(movies, ratings, user_genre){
  # combine rating information into movies
  movieRating = ratings %>% 
    group_by(MovieID) %>% 
    summarize(ratings_per_movie = n(), ave_ratings = round(mean(Rating), dig=3)) %>%
    inner_join(movies, by = 'MovieID')
  
  # Filter movies from selected genre
  mr_selected = movieRating[grep(user_genre, movieRating$Genres),]
  
  # recommender based on popularity
  recommendations = arrange(mr_selected, desc(ave_ratings))
  
  # Select Top 10 Movies
  recommendations = recommendations %>%
    select('MovieID', 'Title', 'ave_ratings', 'ratings_per_movie', 'Genres', 'Year') %>%
    top_n(5, ave_ratings)
  recommendations
}

## Hardcoding Genre as "Thriller"
user_genre = "Thriller"
recom_pop = get_recommendations_by_genre_by_rating(movies, ratings, user_genre)
kable(recom_pop,format = 'markdown')
```

## System 2

### Evaluate Recommender System Models
In System 2 implementation we are building models to evaluate two collaborative recommendation algorithms - user based collaborative filtering algorithm (UBCF) and Singular Value Decomposition Filtering (SVDF). We are using RMSEs as the performance criteria for evaluation.

#### User Based Collaborative Filtering

User-Based Collaborative Filtering approach groups users according to prior usage behavior or according to their preferences, and then recommends an item that a similar user in the same group viewed or liked. 
If user 1 liked movie A, B and C, and if user 2 liked movie A and B, then movie C might make a good recommendation to user 2. The User-Based Collaborative Filtering approach mimics how word-of-mouth recommendations work in real life.

For each new user, these are the steps:

- Measure how similar each user is to the new one. Like IBCF, popular similarity measures are correlation and cosine.
- Identify the most similar users. The options are:
  - Rate the movies rated by the most similar users. The rating is the average rating among similar users and the approaches are:
    - Average rating
    - Weighted average rating, using the similarities as weights
  
- parameter: normalize = 'Z-score', method = 'cosine', nn = 5    
- The parameter *method* is set to *UBCF* .The similarity method used is *cosine* similarity and 5 nearest neighbours are used. 
    
    
#### SVDF (Funk SVD with gradient descend )
- Funk SVD decomposes a matrix (with missing values) into two components U and V. The singular values are folded into these matrices. The approximation for the original matrix can be obtained by R = UV'. 
- This model predicts new data rows by estimating the u vectors using gradient descend and then calculating the reconstructed complete matrix r for these users via r = uV'.
- parameter: normalize = 'Z-score'

#### Read Rating Data Files
```{r}
myurl = "https://liangfgithub.github.io/MovieData/"
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
```


#### Create Train and Test Dataset
```{r}
set.seed(100)
train.id = sample(nrow(ratings), floor(nrow(ratings)) * 0.8)
train = ratings[train.id, ]
#head(train)

test = ratings[-train.id, ]
label = test[c('UserID', 'Rating')]
test$Rating = NA
#head(test)

```

#### Build two Models to predict the Movie Rating For Each User-Movie Pair in the Test Data
#### Create Utility Matrix stored as Sparse Matrix

```{r, eval=FALSE}
models = list(
  UBCF = list(normalize = 'Z-score', method = 'cosine', nn = 5),
  SVDF = list(normalize = 'Z-score')
)
start.time = Sys.time()
R = acast(train, UserID ~ MovieID, value.var = 'Rating')
R = as(R, 'realRatingMatrix')
rmses = rep(0, length(models))
names(rmses) = names(models)


for (m in 1:length(models)) {
  rec = Recommender(R, method = names(models)[m],
      parameter = models[[m]])
  
  recom = predict(rec, R, type = 'ratings')
  rec_list = as(recom, 'list')
  
  # For all lines in test file, one by one
  for (u in 1:nrow(test)){
      userid = as.character(test$UserID[u])
      movieid = as.character(test$MovieID[u])
      
      rating = rec_list[[userid]][movieid]
      test$Rating[u] = ifelse(is.na(rating), 2.5, rating)
  }
  
  rmses[m] = RMSE(test$Rating, label$Rating)
}
end.time = Sys.time()
run.time = as.numeric(difftime(end.time, start.time, units = 'secs'))
```



## RMSE of Models (UBCF and SVDF)
```{r, echo=FALSE, eval=FALSE}
print(rmses)
```

